---
title: "Clustering of Iris Dataset"
author: "SRIHARSHA SURINENI"
date: "April 9, 2017"
output: word_document
---

```{r setup, include=FALSE}
library(arules)
library(installr)
library(arulesViz)
library(fpc)
library(dplyr)
library(reshape2)
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown


```{r cluste analysis}
set.seed(10574454)
food <- read.csv('http://homepages.uc.edu/~maifg/DataMining/data/qry_Food_by_Month.csv')
```

A glimpse of the structure of the data:
```{r cluster analysis of Food_by_month, echo=FALSE}
head(food)
```
##K-means clustering
```{r}
Food<-food[,-c(1,8)]
Food<-scale(Food)
wss<-NA
for (i in 2:20) wss[i] <- sum(kmeans(Food, centers=i)$withinss)
plot(1:20, wss, type="b", xlab="Number of Clusters",ylab="Within groups sum of squares")

```

   From the above plot, it is evident that 8 is a better choice of k, the number of clusters. Alook at how these clusters are separated:
```{r}
fit<-kmeans(Food, 8)
plotcluster(Food, fit$cluster)
```



```{r}
food$cluster<-fit$cluster
itemlist<-food[,c("NickName","cluster")]
itemlist<-itemlist[order(itemlist$cluster),]
write.csv(itemlist,"./itelist.csv", row.names = FALSE)
```
 Here is an example of one of the clusters:

```{r}
cluster5<-food[food$cluster==1,]
cluster5
```

  Mean of the purcases of clusters:

```{r}
means<-aggregate(food[,-1], by = list(food$cluster), mean)
means[,-1]

```

##Hierarchical Clustering:

```{r}
food.dist=dist(Food)
#Obtain clusters using the Wards method
food.hclust=hclust(food.dist, method="ward")

plot(food.hclust)

```

  Capping the nuber of clusters to 8:
  
```{r}
food.8clust = cutree(food.hclust,k=8)

#plotcluster(ZooFood, fit$cluster)

plotcluster(Food, food.8clust, main = "Plot of clusters - Hierarchical clustering")


```

  Here is a look at one of the clusters:

```{r}
food$cluster<-food.8clust
food[food$cluster == 5,]
food[food$cluster == 4,]
```

  Hierarchical clustering gives more coherent clusters with items in same clusers varying in same manner in terms of monthly purchases.
  
    Mean of the purcases of clusters:

```{r}
means<-aggregate(food[,-1], by = list(food$cluster), mean)
means[,-1]

```

