---
title: "Market Basket Analysis of Zoo Dataset"
author: "SRIHARSHA SURINENI"
date: "April 10, 2017"
output: word_document
---

```{r setup, include=FALSE}
library(arules)
library(installr)
library(arulesViz)
library(fpc)
library(dplyr)
library(reshape2)
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

  Here is a glimpse of the structure of the data:
  
```{r cars}
TransFood <- read.csv('http://homepages.uc.edu/~maifg/DataMining/data/food_4_association.csv')
TransFood <- TransFood[, -1]
dat <- as(as.matrix(TransFood), "transactions")

summary(dat)
```

  There are 197 transactions with zero items. Majority of the transactions contain 1 to 5 items. After removing tranactions with zero items as they are of no use in the following analysis, we ave 18879 transactions.
  
  Here is a glimpse of the frequency plot of the items with minimum spport of 0.05:

```{r}
itemFrequencyPlot(dat, support = 0.05, cex.names=0.8)

```
  
  
  
  Applying appriori algorithm to find associaton rules with minimum support of 0.001, confidence of 0.5 and lift ratio of 1.2:
  
```{r, results = 'hide'}
dat = dat[size(dat)>0]

basket_rules <- apriori(dat,parameter = list(sup = 0.001, conf = 0.2, target="rules"))


```

  
```{r}
summary(basket_rules)
inspect(head(basket_rules))
```


  There are very few rules with a good value of support. Majority of the rules have support less than 0.03. This can be expected with the high number of possible combinations of items with varied sizes. Majority of the rules have confidence greater than 0.3, very few of them have confidence greater than 0.5 (which is not a good measure of relaibility) and majority of the rules have lift ratio greater than 4 (which is a good thing, rules are useful). Filtering rules with confidence greater than 0.7.

```{r}
basket1<-subset(basket_rules, confidence>0.7)
summary(basket1)

```

  There are 119 rules with confidence greater than 0.7. Lift ratio of majority f the rules is greater than 7. We lost some rules with better lift ratios than 30 (which is the maximum in this basket of rules). Lets have a look at the rules with lift ratio greater than 20.

```{r}
basket2<-subset(basket_rules, lift>20)

basket3<-subset(basket2, confidence>0.8)
inspect(basket3)

basket4<-subset(basket_rules, lift>60)

```
 
 There are 8 rules with lift ratio greater than 20 and confidence geater than 0.8. Graphical visualisation of these 8 major rules:

```{r}

plot(head(sort(basket3, by="lift")), method = "graph")

```

   These 8 rules are absolutely reliable (confidence approaching 1). Rule with maximum lift ratio (76) is as follows:



```{r}
plot(head(sort(basket4, by="lift")), method = "graph")

```

  Almost all the major rules involve one or the other items of: "Hot Dog Food", "Side of Cheese Food", "Small Drink Food", "Cheese Coney Food" and "Medium Drink Food".



